---
title: "Central Concentration for Nominal Variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Central Concentration for Nominal Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(nomiShape)
```

## Central Concentration for Nominal Variables
This vignette demonstrates how to compute the central concentration for nominal variables using the `central_concentration` function from the `nomiShape` package. Central concentration quantifies how concentrated the distribution of a nominal variable is around its most frequent categories.

### Computing Central Concentration

```{r central-concentration-example-1}
# Example usage of central_concentration
central_concentration(categories, "animal")
```

```{r central-concentration-example-2}
# Example usage of central_concentration
central_concentration(categories2, "animal")
```

```{r central-concentration-example-3}
# Example usage of central_concentration
central_concentration(categories3, "animal")
```

